### 2013-09-26 manuscripts/dpp-msbayes/dpp-msbayes.tex
Working on ABC algorithm description in dpp ms.

\begin{linenomath}
To avoid calculating the likelihood terms of Equation \ref{eq:fullModelCompact},
we distill each sequence alignment \alignment{}{} into a vector of insufficient
summary statistics \alignmentSS{}{}, thus replacing the full dataset
$\alignmentVector = (\alignment{1}{1}, \ldots,
    \alignment{\npairs{}}{\nloci{\npairs{}}})$
with vectors of summary statistics for each alignment
$\ssVector{} = (\alignmentSS{1}{1}, \ldots,
    \alignmentSS{\npairs{}}{\nloci{\npairs{}}})$.
This allows us to estimate the approximate joint posterior distribution
{\small
\begin{equation}
    p(\geneTreeVector, \divTimeMapVector, \demographicParamVector, 
    \locusMutationRateScalarVector, \locusRateHetShapeParameter \given
    \ssSpace, \hkyModelVector, \ploidyScalarVector,
    \mutationRateScalarConstantVector) =
    \frac{p(\ssSpace \given \geneTreeVector, \hkyModelVector)
        p(\geneTreeVector \given \divTimeMapVector, \demographicParamVector,
        \locusMutationRateScalarVector, \ploidyScalarVector,
        \mutationRateScalarConstantVector)
        p(\locusMutationRateScalarVector \given \locusRateHetShapeParameter)
        p(\locusRateHetShapeParameter)
        p(\divTimeMapVector)
        p(\demographicParamVector)}{
        p(\ssSpace)},
    \label{eq:approxModelCompact}
\end{equation}
}
where \ssSpace is the multidimensional Euclidean space around the vector of
summary statistics, the radius of which is the tolerance \tol.
The approximation of this posterior is caused by the insufficiency of the
statistics and the \tol being greater than zero.
We describe the full model in detail before delving into the numerical
method of estimating the approximate model.
\end{linenomath}

@@

\subsection*{The ABC implementation of the model}
To estimate the approximate posterior of Equation \ref{eq:approxModelCompact},
we use an ABC rejection algorithm.
The first step of this algorithm entails collecting a random sample of
parameters values from the joint prior and their associated summary
statistics.
Each sample is generated by
(1) drawing values of all the model's parameters \hpvector{} from their respective
prior distributions,
(2) rescaling the divergence times in \divTimeMapVector from units
proportional to time to units proportional to the expected number
of mutations via Equation X,
(3) simulate a gene tree by drawing coalescent times from
a multi-population Kingman-coalescent model,
(4) simulating sequences along the gene tree under the
\hky substitution parameters of \hkyModelVector,
and (5) calculating population genetic summary statistics from the sequences
\ssVector{} (these are the same statistics calculated from the observed
alignments, \ssVectorObs{}).

>>>>

\begin{linenomath}
To avoid calculating the likelihood terms of Equation \ref{eq:fullModelCompact},
we distill each sequence alignment \alignment{}{} into a vector of insufficient
summary statistics \alignmentSS{}{}, thus replacing the full dataset
$\alignmentVector = (\alignment{1}{1}, \ldots,
    \alignment{\npairs{}}{\nloci{\npairs{}}})$
with vectors of summary statistics for each alignment
$\ssVectorObs = (\alignmentSSObs{1}{1}, \ldots,
    \alignmentSSObs{\npairs{}}{\nloci{\npairs{}}})$.
For each populationn pair, the means of the summary statistics are calculated
across the \nloci{} loci, reducing the vector to
$\ssVectorObs = (\alignmentSSObs{1}{}, \ldots,
    \alignmentSSObs{\npairs{}}{})$.
This allows us to estimate the approximate joint posterior distribution
{\small
\begin{equation}
    p(\geneTreeVector, \divTimeMapVector, \demographicParamVector, 
    \locusMutationRateScalarVector, \locusRateHetShapeParameter \given
    \ssSpace, \hkyModelVector, \ploidyScalarVector,
    \mutationRateScalarConstantVector) =
    \frac{p(\ssSpace \given \geneTreeVector, \hkyModelVector)
        p(\geneTreeVector \given \divTimeMapVector, \demographicParamVector,
        \locusMutationRateScalarVector, \ploidyScalarVector,
        \mutationRateScalarConstantVector)
        p(\locusMutationRateScalarVector \given \locusRateHetShapeParameter)
        p(\locusRateHetShapeParameter)
        p(\divTimeMapVector)
        p(\demographicParamVector)}{
        p(\ssSpace)},
    \label{eq:approxModelCompact}
\end{equation}
}
where \ssSpace is the multidimensional Euclidean space around the vector of
summary statistics, the radius of which is the tolerance \tol.
The approximation of this posterior is caused by the insufficiency of the
statistics and the \tol being greater than zero.
We describe the full model in detail before delving into the numerical
method of estimating the approximate model.
\end{linenomath}

@@

\subsection*{ABC estimation of the posterior}
To estimate the approximate posterior of Equation \ref{eq:approxModelCompact},
we use an ABC rejection algorithm.
The first step of this algorithm entails collecting a random sample of
parameters values from the joint prior and their associated summary
statistics.
Each sample is generated by
(1) drawing values of all the model's parameters, which we denote \hpvector{}, from their respective
prior distributions;
(2) rescaling the divergence times
$\divTimeMapVector = (\divTimeMap{1}, \ldots, \divTimeMap{\npairs{}})$
from units proportional to time to units proportional to the expected number
of mutations via Equation \ref{eq:divTimeScaled} to get 
$\divTimeScaledVector = (\divTimeScaled{1}{1}, \ldots, \divTimeScaled{\npairs{}}{\nloci{\npairs{}}})$;
(3) simulating gene trees $\geneTreeVector = (\geneTree{1}{1}, \ldots, 
\geneTree{\npairs{}}{\nloci{\npairs{}}})$
for each locus of each population pair by drawing coalescent times from
a multi-population Kingman-coalescent model given the demographic parameters;
(4) simulating sequences alignments 
$\alignmentVector = (\alignment{1}{1}, \ldots, \alignment{\npairs{}}{\nloci{\npairs{}}})$
along the gene trees under the
\hky substitution parameters
$\hkyModelVector = (\hkyModel{1}{1}, \ldots, \hkyModel{\npairs{}}{\nloci{\npairs{}}})$
that have the same number of sequences and sequence lengths as the observed
dataset;
(5) calculating population genetic summary statistics
$\ssVector{} = (\alignmentSS{1}{1}, \ldots, \alignmentSS{\npairs{}}{\nloci{\npairs{}}})$
from the simulated sequence alignments (these are the same statistics calculated from the observed
alignments, \ssVectorObs{});
and (6) reducing the summary statistics to the means across loci for each
population pair to get
$\ssVector{} = (\alignmentSS{1}{}, \ldots, \alignmentSS{\npairs{}}{})$, which is
the same summary statstic vector estimated from the observed data \ssVectorObs.
After repeating this procedure \numPriorSamples times, we obtain a random sample
of parameters
from the model prior and there associated summary statistics
.
